---
title: "BRT analyses"
author: "Harald Kliems"
date: "10/21/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

A dude on the internet claimed that the Madison BRT project would "not cater to the bottom 2 deciles of income and those are the groups who typically don't have the option of driving." 

# Get data
## BRT route and stations
Data are available on the City of Madison Open Data portal.
- [stations](https://opendata.arcgis.com/datasets/42118889b9af42358071b0485c53bee6_27.zip)
- [routes](https://opendata.arcgis.com/datasets/42118889b9af42358071b0485c53bee6_27.zip)

For purposes of this analysis, it makes sense to focus on the stations -- a bus that passes by your residence but doesn't stop anywhere nearby doesn't do you any good.

```{r}
library(tidyverse)
library(sf)
library(tidycensus)
```
The dataset contains all stations, including those for planned future lines. I will filter to the ones in the current proposal, the "locally preferred alternative."
```{r}
stations <- read_sf("data/Bus_Rapid_Transit_Stations.shp") %>% 
  filter(LPA == 1)
```

## Population and income data
The American Community Survey has data on income at the block group level. This is a fairly large area for this type of analyses and data at the block level would be preferable. However, only the decennial census provides that level of detail -- which at this point is too old to be useful.

```{r}
vars2018 <- load_variables(2018, "acs5", cache = T)
dp_vars <- load_variables(2018, "acs5/profile", cache = T)
```
Table `B19001` has household income data, providing the number of households in different income brackets. First, we get data for all block groups in Dane County.

```{r}
income <- get_acs("block group", table = "B19001", state = 55, county = "Dane", geometry = T)
```
The median household income for the whole City of Madison is available in the data profile tables. 

```{r}
madison <- get_acs("place", table = "DP03", state = 55) %>% 
  filter(NAME == "Madison city, Wisconsin")

madison %>% filter(variable == "DP03_0062") %>% select(estimate)
```
The median household income in Madison is a little under $63,000. I don't think there is a way to get actual income percentiles from the data.

```{r}

```


# Buffers around stations
To create areas served by BRT stops, as a simple model we create buffers around the stops. How far are people willing to walk to a stop? That's a controversial question. I'll start with the common European spacing of 400 meters.

```{r}
library(units)
buffers <- stations %>% 
  st_transform(3857) %>% 
  st_buffer(set_units(400, metre))
```

```{r}
library(tmap)

tmap_mode("view")
tm_shape(buffers) +
  tm_polygons()
```

